---
title: "Big example"
output: beamer_presentation
date: "2022-11-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, # Required
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      message = FALSE, # Turn off load messages
                      warning = FALSE # Turn off warnings
                      )
```

```{r, include=FALSE}
# Working directories. Uncomment as needed.
# setwd("G:/My Drive/Fall 22/CMDA 4654/proj 2")
setwd("G:/My Drive/Fall 22/CMDA 4654/proj 2")
```


## Data Background

The data was collected using web scrapping sourced from Sports Reference. The 
website includes data collected on different players and sports. Our data in
particular was collected on basketball players from NBA. The data is accurate
up-to the spring semester of 2022.

## Dictionary

Column Name   Definition
-----------   ----------
Name          Player Name
G             Games Played
GS            Games Started
MP            Minutes Played
FG            Field Goals
FGA           Field Goals Attempted
FG2           2 Point Field Goals
FG2A          2 Point Field Goals Attempted
FG3           3 Point Field Goals
FG3A          3 Point Field Goal Attempted
FT            Field Throws
FTA           Field Throws Attempted
ORB           Offensive Rebounds
DRB           Defensive Rebounds

## Dictionary (Continued)

Column Name   Definition
-----------   ----------
TRB           Total Rebounds
AST           Assists Per Game
STL           Steals
BLK           Blocks
TOV           Turnovers
PF            Personal Fouls
PTS           Points
EFG_PCT       Effective Field Goal Percentage
X_PER_G       X Per Game
X_PCT         X Percentage

## What Are We Focusing On?
As you can see there are a lot of variables in this data set and many different analyses can be performed. For now we would like to build a model to predict the number of fouls based on the time each player spends playing.

## Preview
```{r}
players <- read.csv("Data/NBA_players.csv")
head(players)[,c(1,2,3,4,25)]
```

## First Look
```{r, echo=F}
library(ggplot2)
players <- read.csv("Data/NBA_players.csv")
ggplot(players, aes(MP_PER_G, PF_PER_G)) +
  geom_point() + 
  ggtitle("Minutes Played Vs. Fouls") +
  ylab("Personal Fouls Per Game") +
  xlab("Minutes Played Per Game") +
  theme_bw()
```

## Creating a Model
There is a built in function that performs smoothing splines in the 'stats' library. A nice thing about the built-in function is its ability to select the best lambda for us.

```{r}
library(stats)
ss.fit.1 <- 
  smooth.spline(players$MP_PER_G, players$PF_PER_G)
```

If we were doing it "by hand" it would look like this
```{r}
# This will be used for manual model creation.
```


## Smoothing Spline Plot

```{r, echo=F}
ggplot(data = players, mapping = aes(x = MP_PER_G, y = PF_PER_G)) +
  geom_point() +
  geom_line(mapping = 
              aes(x = MP_PER_G, 
                  y = predict(ss.fit.1, players$MP_PER_G)$y,
                  color="red"), 
            show.legend = F, 
            linewidth=1) +
  ggtitle("Minutes Played Vs. Fouls Automatic Smoothing Splines") +
  ylab("Personal Fouls Per Game") +
  xlab("Minutes Played Per Game") +
  theme_bw()

#Add different fits with different lambdas to show effect
```

## Changing our lambda
```{r}
library(stats)
ss.fit.2 <- 
  smooth.spline(players$MP_PER_G, 
                players$PF_PER_G, 
                lambda=0)

ss.fit.3 <- 
  smooth.spline(players$MP_PER_G, 
                players$PF_PER_G, 
                lambda=10000)
```

## Comparing Lambdas
```{r, echo=F}
plt.1 <- ggplot(data = players, mapping = aes(x = MP_PER_G, y = PF_PER_G)) +
  geom_point() +
  geom_line(mapping = 
              aes(x = MP_PER_G, 
                  y = predict(ss.fit.1, players$MP_PER_G)$y,
                  color = "Auto"), show.legend = F, linewidth=1)  + 
    ggtitle("Lambda = Auto") +
    ylab("Personal Fouls Per Game") +
    xlab("Minutes Played Per Game")
  
plt.2 <- ggplot(data = players, mapping = aes(x = MP_PER_G, y = PF_PER_G)) +
  geom_point() +
    geom_line(mapping = 
              aes(x = MP_PER_G, 
                  y = predict(ss.fit.2, players$MP_PER_G)$y,
                  color = "0"), show.legend = F, linewidth=1)  + 
    ggtitle("Lambda = 0") +
    ylab("Personal Fouls Per Game") +
    xlab("Minutes Played Per Game")

plt.3 <- ggplot(data = players, mapping = aes(x = MP_PER_G, y = PF_PER_G)) +
  geom_point() +  
    geom_line(mapping = 
              aes(x = MP_PER_G, 
                  y = predict(ss.fit.3, players$MP_PER_G)$y), color = "red", linewidth=1) + 
    ggtitle("Lambda = 10000 ") +
    ylab("Personal Fouls Per Game") +
    xlab("Minutes Played Per Game")

#Add different fits with different lambdas to show effect
library(gridExtra)
grid.arrange(plt.1, plt.2, plt.3)
```